---
title: "Prctical Machine Learning Project"
output: html_document
Author: Marwan Ismael
---

##Introduction
        
        Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

you can download the Traning & test data sets from below URLs

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv



##Setting environment
Load the required packages and set the seed to a default value

```{r, echo=FALSE, warning=FALSE}
library(caret)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(randomForest)
library(rpart)


set.seed(1984)
```

##Getting the data

Read the Test & Train datasets into memory, taking into consideration files were downloaded earlier.
```{r}
setwd("C:/Users/user/Documents/R/Working Dir")
training <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!",""))
testing <- read.csv("pml-test.csv", na.strings=c("NA","#DIV/0!",""))
```


Split training data set into 2 datasets: Training and Validating (80 % training and 20% validating)
```{r}
inTrain <- createDataPartition(y=training$classe, p=0.8, list=FALSE)
TrainingSub <- training[inTrain, ] 
ValidatingSub <- training[-inTrain, ]
```



##Cleaning the data
Follow below steps to clean the data:

1: identify the near zero variable and remove them:
```{r}
myDataNZV <- nearZeroVar(TrainingSub, saveMetrics=TRUE)
NonZeroVariables <- rownames(myDataNZV[myDataNZV$zeroVar == FALSE,])
TrainingSub <- TrainingSub[NonZeroVariables]
```

2: Remove ID column from Training dataset
```{r}
TrainingSub <- TrainingSub[c(-1)]
```

3: Remove Variables with too many NAs.

```{r}
Temp <- TrainingSub 
for(i in 1:length(TrainingSub)) { 
        if( sum( is.na( TrainingSub[, i] ) ) /nrow(TrainingSub) >= .6 ) { 
        for(j in 1:length(Temp)) {
            if( length( grep(names(TrainingSub[i]), names(Temp)[j]) ) ==1)  { 
                Temp <- Temp[ , -j]
            }   
        } 
    }
}

TrainingSub <- Temp
```

Clean the validate and Test datasets in the same way:

```{r}
clean1 <- colnames(TrainingSub)
ValidatingSub <- ValidatingSub[clean1,]

clean2 <- colnames(TrainingSub[, -58]) 
testing <- testing[clean2,]
```


Convert all variables to same data type
```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(TrainingSub)) {
        if( length( grep(names(TrainingSub[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(TrainingSub[i])
        }      
    }      
}

testing <- rbind(TrainingSub[2, -58] , testing)
testing <- testing[-1,]
```


##Predicting:


Fit Tree model
```{r}
modFit1 <- rpart(classe ~ ., data=TrainingSub, method="class")
```
Plot the tree:

```{r}
fancyRpartPlot(modFit1)
```



predict results using Validating Data Set first:

```{r}
predictionsA1 <- predict(modFit1, ValidatingSub, type = "class")

confusionMatrix(predictionsA1, ValidatingSub$classe)
```

Fit another model using Random Forests and predict errors again to compare with model one.
```{r}
modFitB1 <- randomForest(classe ~. , data=TrainingSub)

predictionsB1 <- predict(modFitB1, ValidatingSub, type = "class")

confusionMatrix(predictionsB1, ValidatingSub$classe)
```



Random Forests accuracy is higher than Tree model

##out of sample error.

Using the Testing data set we can estimate the out of sample error
```{r}
predictionsB2 <- predict(modFitB1, testing, type = "class")
predictionsB2
```



